Collective excitations in competing phases in two and three dimensions

### 2023-12-14 paper/main.tex
stuff

In this article, we will first employ a mean-field approximation to the interaction terms 
and then use the methodology from the so-called iterated equations of motion approach (iEoM),
which has already seen success in the handling of interaction quenches \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
Here, one commutes operators from a suitable operator basis with the Hamiltonian and adds the newly occurring terms to it.
Naturally, a truncation is necessary for most practical applications, however, the approximation becomes better the more terms are included within the basis.
The functionality of this method has been compared to the density matrix formalism and has proved to be considerably more accurate \cite{Kalthoff17}.
Additionally, we will show an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Leggett) and amplitude (Higgs) modes \cite{Cea14,Tsuji15,Krull16,Schwarz20,Fan22}.
The former is located at zero energy as our model describes neutral superfluids.
The latter is found at the lower edge of the two-particle continuum.

The remainder of this paper is organized as follows:
In \autoref{sec:model} we will introduce the model and its Hamiltonian as well as our mean-field theory.
A brief overview of the iterated equations of motion approach is presented in \autoref{sec:ieom}.
Next, we show our results in \autoref{sec:results}.
Lastly, in \autoref{sec:conclusion}, we will discuss our results.
In the appendix \textcolor{red}{(or in the supplemental material?)}, we will derive and explain our methods in detail.

@@

So far we have discussed the static mean-field equations that essentially describe the static ground state.
These results will be used to compute the quantities, namely expectation values, necessary for the iterated equations of motion approach \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
With it, we can describe two-particle quantities like the various collective excitations of the system.
We will present the general ideas briefly here and refer the interested reader to the appendix \textcolor{red}{!REF!} for a detailed explanation and derivation.

@@

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 3000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \textcolor{red}{!REF!}.

@@

Due to this choice of set, the matrices occurring in \eqref{eqn:heisenberg} have a special block structure
\begin{equation}
    \label{eqn:xp_set}
    \mM = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mN = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix}\,,
\end{equation}
where the upper block refers to all $X_i$ and the lower block to all $P_i$.
Furthermore, due to symmetry, the relations $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$ hold.
Thus, if all occurring expectation values are real, both matrices have large empty blocks.
This condition is fulfilled for all cases investigated in this article.
Additionally, the matrix $\mM$ is positive semidefinite if the system is in thermal equilibrium.
We proof this statement in the appendix \textcolor{red}{!REF!}.

@@

Furthermore, assuming, that $\mN$ is invertible, we can easily solve the differential equation \eqref{eqn:heisenberg} to obtain
\begin{equation}
    \vec{c}(t) = \exp \left( i \mN^{-1} \mM t \right) \vec{c}(0)\,.
\end{equation}
This allows us to finally obtain
\begin{align}
    G_{AB^\dagger} (z) &= i \int_0^\infty \exp(izt) \langle [ (\vec{A \cdot} \vec{b}(0))^\dagger, \vec{A} \cdot \vec{a}(t) ] \rangle\ \mathrm{d}t \nonumber \\
        &= i \int_0^\infty \exp(izt) \vec{b}^\dagger(0) \underbrace{\langle [ \vec{A}^\dagger, \vec{A} ] \rangle}_{\equiv \mN} \vec{a}(t) \mathrm{d}t \nonumber \\
        &= i  \int_0^\infty \vec{b}^\dagger(0) \mN \exp \left( i \left(\mN^{-1} \mM + z \right) t \right) \vec{a}(0) \mathrm{d}t  \nonumber \\
        &= - \vec{b}^\dagger(0) \left[ \mN \frac{1}{\mN^{-1} \mM + z} \right] \vec{a}(0) \nonumber \\
        &= - \vec{b}^\dagger(0) \left[ \mN \frac{1}{ \underbrace{\mM + z \mN}}_{\equiv \mathcal{R}(z)} \mN \right] \vec{a}(0)\,,
\end{align}
where $\vec{a}(0)$ and $\vec{b}(0)$ are chosen such, that their product with $\vec{A}$ yields $A(0)$ and $B(0)$, respectively.
The task at hand is finding the resolvent as $\mathcal{R}$.

@@

\appendix
\section{Semipositivity of the dynamical matrix}

@@

\section{Numerical treatment of the iEoM matrices}

We would like to use operators like \eqref{eqn:ieom_basis_operator} to span our operator set.
However, numerically, it is impossible to matrices with infinite dimensions, requiring us to discretize.
Additionally, we cannot numerically write down $\delta$ distributions as matrix elements. Therefore, we need some special treatments, which we will describe in this section.
We discretize the $\gamma$-space equally with each abscissa $\gamma_i$ being spaced by $\Delta \gamma$.
Then, we define an approximate $\delta$ function
\begin{equation}
    h(\gamma) \coloneqq \begin{cases}
        \frac{1}{\Delta \gamma} & |\gamma| < \frac{\Delta \gamma}{2} \\ 0 & \text{otherwise}
    \end{cases}\,.
\end{equation}
Computing an integral yields
\begin{equation}
    \int f(\tilde{\gamma}) h(\tilde{\gamma} - \gamma_j) \mathrm{d} \tilde{\gamma} \approx \sum_i f(\gamma_i) h(\gamma_i - \gamma_j) \Delta \gamma = f(\gamma_j)\,,
\end{equation}
and thereby mimicking the continuous $\delta$ distribution for a discrete description.
Note, that if one takes the limit $\Delta \gamma \to 0$ there $\delta$ distribution is exactly reproduced.

Numerically, we can compute the matrix elements like this
\begin{align}
    (A_i | A_j) &= \frac{(\Delta \gamma)^2}{N} \sum_{\vk \vl} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vl)) (A_{\vk} | A_{\vl})\,.
\end{align}
Here, all expressions are finite numbers and do not require further treatment.
Mutatis mutandis, one obtains a similar expression for $(A_i | [H, A_j])$.
Additionally, note, that the expressions $(A_{\vk} | A_{\vl})$ and $(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums of terms that are proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of momentum space computations.

>>>>

In this article, we will first employ a mean-field approximation to the interaction terms 
and then use the methodology from the so-called iterated equations of motion approach (iEoM),
which has already seen success in the handling of interaction quenches \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
Here, one commutes operators from a suitable operator basis with the Hamiltonian and adds the newly occurring terms to it.
Naturally, a truncation is necessary for most practical applications, however, the approximation becomes better the more terms are included within the basis.
The functionality of this method has been compared to the density matrix formalism and has proved to be considerably more accurate \cite{Kalthoff17}.
Additionally, we will show an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Leggett) and amplitude (Higgs) modes \cite{Varma02,Cea14,Measson14,Tsuji15,Krull16,Schwarz20,Fan22}.
The former is located at zero energy as our model describes neutral superfluids.
The latter is found at the lower edge of the two-particle continuum.

The remainder of this paper is organized as follows:
In \autoref{sec:model} we will introduce the model and its Hamiltonian as well as our mean-field theory.
A brief overview of the iterated equations of motion approach is presented in \autoref{sec:ieom}.
Next, we show our results in \autoref{sec:results}.
Lastly, in \autoref{sec:conclusion}, we will discuss our results.
%In the appendix \textcolor{red}{(or in the supplemental material?)}, we will derive and explain our methods in detail.

@@

So far we have discussed the static mean-field equations that essentially describe the static ground state.
These results will be used to compute the quantities, namely expectation values, necessary for the iterated equations of motion approach \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
With it, we can describe two-particle quantities like the various collective excitations of the system.

@@

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 3000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \autoref{sec:numerical_ieom}.

@@

Due to this choice of set, the matrices occurring in \eqref{eqn:heisenberg} have a special block structure
\begin{equation}
    \label{eqn:xp_set}
    \mM = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mN = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix}\,,
\end{equation}
where the upper block refers to all $X_i$ and the lower block to all $P_i$.
Furthermore, due to symmetry, the relations $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$ hold.
Thus, if all occurring expectation values are real, both matrices have large empty blocks.
This condition is fulfilled for all cases investigated in this article.
Additionally, the matrix $\mM$ is positive semidefinite if the system is in thermal equilibrium.
We proof this statement in the appendix \autoref{sec:positive_M}.

@@

Furthermore, assuming, that $\mN$ is invertible, we can easily solve the differential equation \eqref{eqn:heisenberg} to obtain
\begin{equation}
    \vec{c}(t) = \exp \left( i \mN^{-1} \mM t \right) \vec{c}(0)\,.
\end{equation}
This allows us to finally obtain
\begin{align}
    \label{eqn:green_derivation}
    G_{AB^\dagger} (z) &= i \int_0^\infty \exp(izt) \langle [ (\vec{A \cdot} \vec{b}(0))^\dagger, \vec{A} \cdot \vec{a}(t) ] \rangle\ \mathrm{d}t \nonumber \\
        &= i \int_0^\infty \exp(izt) \vec{b}^\dagger(0) \underbrace{\langle [ \vec{A}^\dagger, \vec{A} ] \rangle}_{\equiv \mN} \vec{a}(t) \mathrm{d}t \nonumber \\
        &= i  \int_0^\infty \vec{b}^\dagger(0) \mN \exp \left( i \left(\mN^{-1} \mM + z \right) t \right) \vec{a}(0) \mathrm{d}t  \nonumber \\
        &= - \vec{b}^\dagger(0) \left[ \mN \frac{1}{\mN^{-1} \mM + z} \right] \vec{a}(0) \nonumber \\
        &= - \vec{b}^\dagger(0) \left[ \mN \frac{1}{ \underbrace{\mM + z \mN}}_{\equiv \mathcal{R}(z)} \mN \right] \vec{a}(0)\,,
\end{align}
where $\vec{a}(0)$ and $\vec{b}(0)$ are chosen such, that their product with $\vec{A}$ yields $A(0)$ and $B(0)$, respectively.
The task at hand is finding the resolvent as $\mathcal{R}$.

@@

\appendix
\section{Semipositivity of the dynamical matrix}
\label{sec:positive_M}

@@

\section{Numerical treatment of the iEoM matrices}
\label{sec:numerical_ieom}

We would like to use operators like \eqref{eqn:ieom_basis_operator} to span our operator set.
However, numerically, it is impossible to matrices with infinite dimensions, requiring us to discretize.
Additionally, we cannot numerically write down $\delta$ distributions as matrix elements. Therefore, we need some special treatments, which we will describe in this section.
We discretize the $\gamma$-space equally with each abscissa $\gamma_i$ being spaced by $\Delta \gamma$.
Then, we define an approximate $\delta$ function
\begin{equation}
    h(\gamma) \coloneqq \begin{cases}
        \frac{1}{\Delta \gamma} & |\gamma| < \frac{\Delta \gamma}{2} \\ 0 & \text{otherwise}
    \end{cases}\,.
\end{equation}
Computing an integral yields
\begin{equation}
    \int f(\tilde{\gamma}) h(\tilde{\gamma} - \gamma_j) \mathrm{d} \tilde{\gamma} \approx \sum_i f(\gamma_i) h(\gamma_i - \gamma_j) \Delta \gamma = f(\gamma_j)\,,
\end{equation}
showing that $h(\gamma)$ mimicks the continuous $\delta$ distribution for a discrete description.
Note, that if one takes the limit $\Delta \gamma \to 0$ the $\delta$ distribution is exactly reproduced.

Numerically, we can compute the matrix elements like this
\begin{equation}
    \label{eqn:numerical_example}
    (A_i | A_j) = \frac{1}{N} \sum_{\vk \vl} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vl)) (A_{\vk} | A_{\vl})\,.
\end{equation}
Here, all expressions are finite numbers and do not require further treatment.
Mutatis mutandis, one obtains a similar expression for $(A_i | [H, A_j])$.
Additionally, note, that the expressions $(A_{\vk} | A_{\vl})$ and $(A_{\vk} | [H, A_{\vl}])$ can be reduced to sums of terms that are proportional to $\gamma_i$ or $\gamma_j$, respectively.
Therefore, it is not necessary to implement any sort of momentum space computations.
For example, consider $A_{\vk} = f_{\vk}^\dagger$, then
\begin{equation}
    ( f_{\vk}^\dagger | f_{\vl}^\dagger ) = \delta_{\vk, \vl} \langle 1 - n_{\vk \uparrow} - n_{\vk \downarrow} \rangle \eqqcolon \delta_{\vk, \vl} \alpha(\vk)
\end{equation}
holds. Inserting this into \eqref{eqn:numerical_example} yields
\begin{align}
    (A_i | A_j) &= \frac{1}{N} \sum_{\vk} h(\gamma_i - \hat{\gamma}(\vk)) h(\gamma_j - \hat{\gamma}(\vk)) \alpha(\vk)\,.
\end{align}
As stated in the main text, see \eqref{eqn:equal_expecs}, $\alpha(\vk)$ depends only on $\hat{\gamma}(\vk)$, hence we can switch from the momentum sum to a $\gamma$ integration
\begin{align}
    (A_i | A_j) &= \int \rho(\gamma) h(\gamma_i - \gamma) h(\gamma_j - \gamma) \alpha(\gamma) \dgamma \nonumber \\
        &\approx \rho(\gamma_i) h(\gamma_j - \gamma_i) \alpha(\gamma_i) \nonumber \\
        &= \frac{\delta_{ij}}{\Delta \gamma} \rho(\gamma_i) \alpha(\gamma_i)\,.
\end{align}
Later on, when we compute the spectral functions, we perform another two $\gamma$ integrations, for example, let us compute the first Lanczos coefficient for $A_\text{Higgs}$.
Then, the elements of the vectors mentioned in \eqref{eqn:green_derivation} are given by
\begin{equation}
    a_i = b_i = \begin{cases}
        \Delta \gamma & i \text{ corresponds to an } f\text{-type basis operator} \\
        0 & \text{otherwise}
    \end{cases}\,.
\end{equation}
